---
title: "Data Exploration"
author: "Heru Handika"
date: "2/16/2021"
output: html_document
---

```{r}
if (!require(pacman)) {
  install.packages(pacman)
}

pacman::p_load(here)
pacman::p_load(infer)
pacman::p_load(tibble)
pacman::p_load(tidyr)
pacman::p_load(dplyr)
pacman::p_load(ggplot2)


source(here("R", "colors.R"))
source(here("R", "utils.R"))

package_list <- c("here", "pacman", "infer", "tidyr", "dplyr", "ggplot2", "tibble")

generate_citation <- function(packageName) {
  packageVersion(packageName)
  citation(packageName)
}

purrr::map(package_list, generate_citation)

RStudio.Version()
```

## Dataset selection

Here, I use the morphology data as source of files. Then I select 20 skull variables to create a dataset for skull data analyses. 


```{r}
morph_all <- readr::read_csv(here("data", "morphology", "morpho_all.csv"))

ext <- morph_all |> 
  dplyr::select(species, voucher_no, site_id, HB, TL, HF, ER, Mass)

skulls <- morph_all |> 
  dplyr::select(voucher_no, species, GLS:DUI)

species_group2 <- c("Bunomys penitus", "Frateromys fratrorum", "Frateromys spnov", "Lenomys meyeri", "Taeromys dominator")

species_group1 <- c("Bunomys chrysocomus", "Frateromys fratrorum", "Frateromys spnov", "Taeromys dominator", "Taeromys taerae")
```

## Test statistics for sexual dimorphism

```{r}
frateromys.df <- morph_all |> 
  tidyr::drop_na(GLS) |> 
  dplyr::filter(sex != "Unknown") |>
  dplyr::filter(genus == "Frateromys")

temboan.df <- frateromys.df |>
  dplyr::filter(site_id == "Temboan") |> 
  dplyr::filter(sex != "n/a")

temboan.df |> 
  group_by(sex) |> 
  dplyr::summarise(N = n())

temboan.df |> 
  infer::t_test(response = GLS, explanatory = sex)

count <- frateromys.df |> 
  dplyr::group_by(species, site_id, sex) |> 
  dplyr::count()

count
```

## Summarize external characters


```{r}
ext_sum <- ext |>
  dplyr::group_by(species) |>
  dplyr::select(-voucher_no)

sum <- ext_sum |> 
  dplyr::summarize_all(tibble::lst(mean, min, max, sd))

HB_stats <- ext_sum |>
  tidyr::drop_na() |>
  dplyr::summarise(N = n(), HB_mean = mean(HB), HB_min = min(HB), HB_max = max(HB), HB_sd = sd(HB))

TL_stats <- ext_sum |>
  tidyr::drop_na() |>
  dplyr::summarise(N = n(), TL_mean = mean(TL), TL_min = min(TL), TL_max = max(TL), TL_sd = sd(TL))

ext |>
  dplyr::group_by(species) |>
  dplyr::summarise(n())

sample_per_site <- ext |>
  dplyr::group_by(species) |>
  dplyr::count(site_id)

sum_per_site <- ext |>
  dplyr::select(species, site_id, HB, TL) |>
  tidyr::drop_na() |>
  dplyr::mutate(hb_tl = TL / HB * 100) |>
  dplyr::group_by(species, site_id) |>
  dplyr::summarise(N = n(), HB_TL_mean = mean(hb_tl), TL_mean = mean(TL), TL_min = min(TL), TL_max = max(TL), TL_sd = sd(TL))
```

## Box plot

```{r}
ext |> 
  dplyr::select(species, HB:ER) |> 
  dplyr::filter(species %in% target) |> 
  tidyr::drop_na() |> 
  tidyr::gather(key = "ext_char", value = "measurement", HB:ER) |>
  dplyr::mutate(ext_char = fct_reorder(ext_char, measurement, .desc = TRUE)) |>
  ggplot(aes(species, measurement)) + 
  geom_boxplot(aes(fill = species)) +
  facet_wrap(~ext_char, ncol = 2, scale="free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) + 
  labs(x = "Species", y = "Measurements")
  

save_plot(fname = "ext_char.pdf", .width = 8, .height = 5)
```
## Shared plot function
This function is used for creating scatter plots across the downstream analyses.

```{r}
plot <- function(plot_data) {
  plot_data +
    geom_point(aes(color = species, shape = species), size = 3) +
    scale_color_drunken_mice() +
    scale_shape_manual(values = c(1, 15, 16, 2, 0, 5)) +
    theme_classic(base_size = 12) +
    theme(legend.position = "right")
}
```


### Scatter plot for External Character

```{r}
ext_p <- ggplot(ext_scatter_df, aes(HB, TL))

p <- plot(ext_p) + labs(x = "Head and Body Length (mm)", y = "Tail Length (mm)")
p
save_ext("hb_tl_scatterplot.pdf")
```

## PCA for Skulls

```{r}
skull_clean <- skulls |>
  drop_na() |>
  filter(species %in% target)

skull_pca <- prcomp(skull_clean[, 3:22], center = TRUE)
```

### Inspect PCA

```{r}
eigen <- skull_pca$sdev^2
eigen

loadings <- skull_pca$rotation
loadings

summary(skull_pca)$importance
```

### Plot PCA

```{r}
skull_df <- tibble(species = skull_clean$species, PC1 = skull_pca$x[, 1], PC2 = skull_pca$x[, 2], PC3= skull_pca$x[, 3])

pca_12 <- ggplot(skull_df, aes(PC1, PC2))
pca_13 <- ggplot(skull_df, aes(PC1, PC3))

p2 <- plot(pca_12) + labs(x = "PC1 (89.7%)", y = "PC2 (2.2%)")
save_skull("pca_20skulls_pc1_pc2.pdf")


p3 <- plot(pca_13) +labs(x = "PC1 (89.7%)", y = "PC3 (1.4%)")
save_skull("pca_20skulls_pc1_pc3.pdf")

```
