---
title: "Data Exploration"
author: "Heru Handika"
date: "2/16/2021"
output: html_document
---

```{r}
if (!require(pacman)) {
  install.packages(pacman)
}

pacman::p_load(here)
pacman::p_load(infer)
pacman::p_load(forcats)
pacman::p_load(tibble)
pacman::p_load(tidyr)
pacman::p_load(dplyr)
pacman::p_load(ggplot2)
pacman::p_load(readxl)
pacman::p_load(ggpubr)
pacman::p_load(factoextra)


source(here("R", "colors.R"))
source(here("R", "utils.R"))
```

## Dataset selection

Load and subset datasets for external and skull morphology analyses.

```{r}
morph_all <- readxl::read_excel(here("data", "morphology", "morpho_all.xlsx")) |> 
  dplyr::mutate(species = stringr::str_replace(species, pattern = "Frateromys spnov", "Frateromys sororum")) |> 
  dplyr::mutate(site_id = stringr::str_replace(site_id, pattern = "Mt. Dua Saudara", "Dua Saudara"))

abbreviate_genus <- function(name) {
  name %>%
    stringr::str_replace("^([A-Za-z])\\w+\\s", "\\1. ")
}

ext <- morph_all |> 
  dplyr::mutate(species = abbreviate_genus(species)) |>
  dplyr::select(species, voucher_no, site_id, HB, TL, HF, ER, WT) 

skulls <- morph_all |> 
  dplyr::mutate(species = abbreviate_genus(species)) |>
  dplyr::select(voucher_no, species, site_id, GLS:DUI)

combined <- morph_all |> 
  dplyr::select(voucher_no, species, HB:DUI)
```
## Test statistics for sexual dimorphism

```{r}
frateromys.df <- morph_all |> 
  tidyr::drop_na(GLS) |> 
  dplyr::filter(sex != "Unknown") |>
  dplyr::filter(genus == "Frateromys")

temboan.df <- frateromys.df |>
  dplyr::filter(site_id == "Temboan") |> 
  dplyr::filter(sex != "n/a")

temboan.df |> 
  group_by(sex) |> 
  dplyr::summarise(N = n())

temboan.df |> 
  infer::t_test(response = GLS, explanatory = sex)

count <- frateromys.df |> 
  dplyr::group_by(species, site_id, sex) |> 
  dplyr::count()

count
```

## Summarize external characters


```{r}
ext_sum <- ext |>
  dplyr::group_by(species) |>
  dplyr::select(-voucher_no)

sum <- ext_sum |> 
  dplyr::summarize_all(tibble::lst(mean, min, max, sd))

HB_stats <- ext_sum |>
  tidyr::drop_na() |>
  dplyr::summarise(N = n(), HB_mean = mean(HB), HB_min = min(HB), HB_max = max(HB), HB_sd = sd(HB))

TL_stats <- ext_sum |>
  tidyr::drop_na() |>
  dplyr::summarise(N = n(), TL_mean = mean(TL), TL_min = min(TL), TL_max = max(TL), TL_sd = sd(TL))

ext |>
  dplyr::group_by(species) |>
  dplyr::summarise(n())

sample_per_site <- ext |>
  dplyr::group_by(species) |>
  dplyr::count(site_id)

sum_per_site <- ext |>
  dplyr::select(species, site_id, HB, TL) |>
  tidyr::drop_na() |>
  dplyr::mutate(hb_tl = TL / HB * 100) |>
  dplyr::group_by(species, site_id) |>
  dplyr::summarise(N = n(), HB_TL_mean = mean(hb_tl), TL_mean = mean(TL), TL_min = min(TL), TL_max = max(TL), TL_sd = sd(TL))
```

## Box plot

```{r}
species_group1 <- c("F. fratrorum", "F. sororum", "B. chrysocomus", "T. dominator", "T. taerae")
species_group2 <- c("B. penitus", "F. fratrorum", "F. sororum", "L. meyeri", "T. dominator")
var_levels <- c("HB", "TL", "HF", "ER", "WT", "Percent TL/HB")

p1 <- ext |> 
  dplyr::select(species, HB:WT) |> 
  dplyr::filter(species %in% species_group1) |>
  dplyr::mutate(species = forcats::fct_relevel(species, species_group1)) |> 
  tidyr::drop_na() |>
  dplyr::mutate("Percent TL/HB" = WT/HB * 100) |> 
  pivot_longer(cols = HB:"Percent TL/HB", names_to = "ext_char", values_to = "measurement") |>
  dplyr::mutate(ext_char = forcats::fct_relevel(ext_char, var_levels)) |>
  ggplot(aes(species, measurement)) + 
    geom_boxplot(aes(fill = species), alpha = 0.8) +
    facet_wrap(vars(ext_char), ncol = 2, scale="free_y") +
    scale_fill_brewer(palette = "Dark2") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 50, hjust = 1, face = "italic"), legend.position="none") + 
    labs(x = "Species", y = "Measurements")
  
p1
save_plot(fname = "ext_char.pdf", .width = 8, .height = 5)
```


## PCA for Skulls

```{r}
skull_clean <- skulls |>
  tidyr::drop_na() |>
  dplyr::filter(species %in% species_group1)

skull_pca <- prcomp(skull_clean[,4:23])
```

### Inspect variable contribution

```{r}
factoextra::fviz_pca_var(
  skull_pca,
  col.var = "contrib", # Color by contributions to the PC
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE     # Avoid text overlapping
)

factoextra::fviz_contrib(skull_pca, choice = "var", axes = 1, top = 10)
factoextra::fviz_contrib(skull_pca, choice = "var", axes = 2, top = 10)
factoextra::fviz_contrib(skull_pca, choice = "var", axes = 3, top = 10)
```


### Inspect PCA

```{r}
eigen <- skull_pca$sdev^2 |> 
  tibble::as_tibble(rownames = "PC") 
skull_pca

loadings <- skull_pca$rotation 

loadings.df <- loadings |> 
  tibble::as_tibble(rownames = "Variables") |> 
  dplyr::select(Variables, PC1, PC2, PC3)

pull_eigen <- function(pc) {
    value <- eigen |> 
      filter(PC == pc) |> 
      pull(value)
    value
}

eigen.df <- tibble(
  Variables = "Eigen value", 
  PC1 = pull_eigen(1),
  PC2 = pull_eigen(2),
  PC3 =  pull_eigen(3))

summary <- summary(skull_pca)$importance |> 
  tibble::as_tibble(rownames = "Variables") |> 
  dplyr::select(Variables, PC1, PC2, PC3)

pull_variance <- function(pc) {
  value <- summary |> 
    filter(Variables == "Proportion of Variance") |> 
    pull(pc)
  round(value * 100, digits = 1)
}

prop_pc1 <- pull_variance("PC1")
prop_pc2 <- pull_variance("PC2")
prop_pc3 <- pull_variance("PC3")

summary.df <- loadings.df |> 
  tibble::add_row(eigen.df) |> 
  tibble::add_row(summary) |> 
  dplyr::filter(Variables != "Cumulative Proportion")

readr::write_csv(summary.df, here::here("results", "loadings.csv"))
```
## Shared plot function

This function is used for creating scatter plots across the downstream analyses.

```{r}
plot <- function(plot_data) {
  plot_data +
    geom_point(aes(color = species), alpha = 0.8, size = 3) +
    # scale_shape_manual(values = c(1, 15, 16, 2, 0, 5)) +
    scale_color_brewer(palette = "Dark2", name = "Species") +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom", legend.text = element_text(size = 10, face = "italic")) +
    guides(col = guide_legend(ncol = 3))
}
```

### Plot PCA

```{r}
skull_df <- tibble(
  species = skull_clean$species, 
  sites = skull_clean$site_id, 
  PC1 = skull_pca$x[, 1], 
  PC2 = skull_pca$x[, 2], 
  PC3= skull_pca$x[, 3]
  ) |> 
  dplyr::mutate(species = forcats::fct_relevel(species, species_group1)) 

pca_12 <- ggplot(skull_df, aes(PC1, PC2))
pca_13 <- ggplot(skull_df, aes(PC1, PC3))

create_axis_title <- function(pc, value) {
  paste0("PC", pc, "(", value, "%)")
}

p2 <- plot(pca_12) + labs(x = create_axis_title(1, prop_pc1), y = create_axis_title(2, prop_pc2))
save_plot(fname = "pca_20skulls_pc1_pc2.pdf")


p3 <- plot(pca_13) + labs(x = create_axis_title(1, prop_pc1), y = create_axis_title(3, prop_pc3))
save_plot(fname = "pca_20skulls_pc1_pc3.pdf")

frateromys <- c("F. fratrorum", "F. sororum")

fratromys.plot <- skull_df |> 
  dplyr::filter(species %in% frateromys) |>
  ggplot(aes(PC1, PC3, shape = sites)) +
    geom_point(aes(color = species), alpha = 0.8, size = 3) +
    scale_shape_manual(values = c(1, 2, 16, 15, 17, 5, 6, 5), name = "Sites") +
    scale_color_brewer(palette = "Dark2", name = "Species") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", legend.text = element_text(size = 10, face = "italic")) +
    guides(col = guide_legend(ncol = 3)) +
    labs(x = create_axis_title(1, prop_pc1), y = create_axis_title(3, prop_pc3))

save_plot(fname = "frateromys_pops_pc13.pdf")
```
## Final plots with layout

```{r}
p23 <- ggpubr::ggarrange(p2, p3, ncol = 2, labels = c("B", "C"), common.legend = TRUE, legend = "bottom")
ggpubr::ggarrange(p1, p23, fratromys.plot, nrow = 3, labels = c("A", "", "D"))

save_plot(fname = "figure_combined.pdf", .width = 8, .height = 12)
```

## Citation

```{r}
package_list <- c("here", "pacman", "infer", "tidyr", "dplyr", "ggplot2", "tibble")

generate_citation <- function(packageName) {
  packageVersion(packageName)
  citation(packageName)
}

purrr::map(package_list, generate_citation)

RStudio.Version()
```

